---
title: "PA1_template"
author: "HOWASAM"
date: "August 14, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
##knitr::opts_knit$set(root.dir = "C:Users/howasam/Desktop/AdHoc/Coursera/Reproduciable Resarch/Lesson 2")
```

## Loading and preprocessing the data


The only prep work required for this data with exception of handling nulls which occurs latter in the assignment is to read in the data from a CSV. This section also installs the package required for latter plotting.

```{r}
library(lattice)

rawData <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day

1) Calculate the total number of steps taken per day

  This code groups the step totals by day and then renames the columns for better readability

```{r}
  aggDate <- aggregate(rawData$steps, by = list(rawData$date), FUN = sum)
  colnames(aggDate)<-c("date" , "steps")
  
```
2) Make a histogram of the total number of steps taken each day

```{r}
  hist(aggDate$steps)
```


3) Calculate and report the mean and median of the total number of steps taken per day

```{r}  
##Mean
  mean(aggDate$steps,na.rm = TRUE)
##Median
  median(aggDate$steps,na.rm  =TRUE)
```

## What is the average daily activity pattern?

1) Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
  This code groups steps into 5 minute intervals and calculates the average and then plots it on a line plot
  
```{r}
 aggInterval <- aggregate(rawData$steps, by = list(rawData$interval), FUN = mean,na.rm = TRUE)
  
  colnames(aggInterval)<-c("interval" , "steps")
  
  ##LinePlot
  plot(x = aggInterval$interval,y = aggInterval$steps, type = 'l' )
```

2) Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?
```{r}
  ##Interval With Max Avg steps
  maxInt = subset(aggInterval, steps== max(steps))
  maxInt$interval
```

##Imputing missing values

1) Calculate and report the total number of missing values in the data set (i.e. the total number of rows with NAs)
```{r}
 ##Number of NA rows
  nrow(subset(rawData,is.na(rawData$steps)==TRUE))
```

2) Devise a strategy for filling in all of the missing values in the data set. 
  In this code I chose to fill in missing values with the average for that 5 minute interval.

3) Create a new data set that is equal to the original data set but with the missing data filled in.
```{r}  
  ##FIll in NA
  
  fillData <- rawData
  
  fillData = merge(x=fillData, y = aggInterval,by.x  = c('interval'),by.y  = c('interval'),all.X = TRUE)
  
  colnames(fillData)<-c("interval" , "steps","date","AvgSteps")
  
  ##Data filled in using the average for that interval 
  fillData$steps[is.na(fillData$steps)]<-fillData$AvgSteps[is.na(fillData$steps)]
```
4) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}   
  aggDateFill <- aggregate(fillData$steps, by = list(fillData$date), FUN = sum)
  
  
  colnames(aggDateFill)<-c("date" , "steps")
  
  
  hist(aggDateFill$steps)
  
  mean(aggDateFill$steps,na.rm = TRUE)
  median(aggDateFill$steps,na.rm  =TRUE)
```

The mean changes not at all because I have used an average value to fill in the missing amounts so it counteracts itself. The Median moved to the mean because there were more instances of that value appearing then used to be present. mitigating outliers

##Are there differences in activity patterns between weekdays and weekends?


1) Create a new factor variable in the data set with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
 ##FIll data by day of week
  fillData$doW<-weekdays(as.Date(fillData$date))
  
  weekDaysList <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
  
  fillData$weekday = factor((fillData$doW %in% weekDaysList), levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
  
  
  
  aggFillInterval <- aggregate(fillData$steps, by = list(fillData$interval,fillData$weekday), FUN = mean,na.rm = TRUE)
  
  colnames(aggFillInterval)<-c("interval" ,"weekday", "steps")

```
Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}    
  ##LinePlot
  with(aggFillInterval, xyplot(steps~interval | weekday, type = 'l' , layout = c(1,2,1) ))
  ```